Whole-body connectome of a segmented annelid larva

### 2024-03-16 Manuscript.qmd
add content to public repo

---
title: "Whole-body connectome of a segmented annelid larva"
author: 
  - name: "Gáspár Jékely<sup>1,2,*</sup>, Csaba Verasztó<sup>2,3</sup>, Sanja Jasek<sup>1,2</sup>, Martin Gühmann<sup>4</sup>, Luis Alberto Bezares Calderón<sup>2</sup>, Elizabeth A. Williams<sup>2</sup>, Réza Shahidi<sup>2,5</sup>"
    affiliation: " <sup>1</sup>Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany <br>  <sup>2</sup>Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom <br>  <sup>3</sup>École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland <br> <sup>4</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom <br> <sup>5</sup>Electron Microscopy Core Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany <br>  <sup>*</sup>Correspondence: gaspar.jekely @ cos.uni-heidelberg.de"
output:
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
  html_document:
    theme: lumen
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

We previously reported synaptic connectomes for several whole-body circuits from a transmission EM  volume of a three-day-old larva [@Randel2015].
These include the visual, startle, ciliomotor, nuchal organ, and neurosecretory systems [@bezares-calderon2018; @Randel2015; @shahidi2015serial; @veraszto2017; @veraszto2018; @williams2017].
Here, we report the complete synaptic connectome and the cell-type complement from the same volume of a three-day-old *Platynereis* larva.
Recently we also reported the desmosomal connectome and motoneuron innervation of the somatic musculature [@jasek2022].
The full connectome reconstruction has now allowed us to uncover several new circuits and to consider all circuits in a whole-body context.
We also found several neurons that span the entire length of the larva, highlighting the strength of a whole-body dataset.
These cells and their circuits allow us to generate hypotheses on how whole-body coordination may be achieved by the larval nervous system.
The connectome also allows us to address long-standing hypotheses about the origin of the segmented annelid body-plan and explore patterns of circuit evolution.

@@

```{r, fig.cap="**Figure 1. Overview of nervous-system anatomy in the three-day-old *Platynereis* larva.** (A) Stylised scanning electron microscopy image of a three-day-old segmented *Platynereis* larva. (B) Morphological rendering of all cells in the EM volume. Spheres represent the position of cell somas. (C) Neurite processes of all neurons in the larva coloured by neuron type (ventral view). Soma are not shown. (D) All effector cells in the larva, including ciliated cells (yellow), glands (grey) and muscles (red). Spheres indicate cell soma (not shown for muscle cells) (E) Summary of cell numbers of different categories in the larva.  (F-H) All presynaptic sites coloured by neuron type shown in (F) ventral, lateral (G), and frontal (H) views. (I) Morphological rendering of all sensory and interneurons on the left side of the larva.  (J) Morphological rendering of all sensory and motor neurons on the left side of the larva. Horizontal lines indicate the position of the cross-sections in (K). (K) Cross-section view of sensory, inter- and motor neuron projections in the ventral nerve cord neuropil. Cross-sections at three antero-posterior positions are shown. Numbers indicate the position of the cross-sections as marked in (J). (L) Position of presynaptic and postsynaptic sites on all sensory neurons of the connectome. (M) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in sensory neurons. (N) Position of presynaptic and postsynaptic sites on all interneurons of the connectome. (O) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in interneurons. (P) Position of presynaptic and postsynaptic sites on all motor neurons of the connectome. (Q) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in motor neurons. In all anatomical renderings, the outline of the yolk is shown in grey. In (G), the body outline is also shown. Aciculae and chaetae are also shown in grey as segmental markers. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron, sg0-3, segments 0-3. <br> Figure 1—source data 1. Source data for panels M, O, Q. \n\n"}
#| echo: false 
knitr::include_graphics(path = "Figures/Figure1.png")
```

@@

To identify more strongly connected subgraphs within the connectome, we used community detection with the Leiden algorithm [@traag2019louvain].
This analysis combined with force-field-based clustering highlighted several modules (Figures 2).
Nodes in a module are more strongly interconnected than to nodes in other modules (Figure 2---figure supplement 3).
We named the modules based on their primary effector organs or other dominant anatomical characters. We would like to note that this analysis can recover different number of modules depending on the resolution parameter (Figure 2—figure supplement 2C). The selected parameters results in communities that we think best represent functional units.

@@

The modules are also interconnected among themselves suggesting crosstalk. Most sensory neurons occur in the visual, the two mechanosensory, the mushroom body and the anterior neurosecretory modules (Figure 2---figure supplement 1J). In a Sankey diagram representing information flow, these modules are upstream of the motor-dominated pigmentmotor, ciliomotor, and musclemotor modules (Figure 2---figure supplement 2B).

```{r, fig.cap="**Figure 2. Modularity of the *Platynereis* larval connectome.** Full connectome graph of the *P. dumerilii* three-day-old larva chemical synapse connectome. Nodes represent individual cells, edges represent synaptic connectivity. Nodes are coloured by modules. Node sizes are proportional to weighted degree. The individual panels show the morphology of all cells in each module with the outline of the yolk and aciculae shown in grey. <br> Figure 2—source data 1. Source data of the connectome graph in tibble graph (tbl_graph) format saved as an R binary object. \n\n"}
#| echo: false 
knitr::include_graphics("Figures/Figure2.png")
```

@@

```{r, fig.cap="**Figure 3. Cell-type complement of the three-day-old *Platynereis* larva.** Segmental distribution and number of sensory neuron types (A), interneuron types (B) and motoneuron types (C). Histogram of the number of cells per neuronal (D) and non-neuronal (E) cell types. (F) Segmental distribution and number of non-neuronal cell types. The number in each box refers to the number of cells of the indicated cell type in the indicated segment. <br> Figure 3—source data 1. Source data for panels A-C. <br> Figure 3—source data 2. Source data for panel F. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure3.png")
```

@@

```{r, fig.cap="**Figure 4. Cell-type-level connectivity of the *Platynereis* larval connectome.** (A) The cell-type connectome of the three-day-old larva. Nodes represent grouped cells of the same type, edges represent synaptic connectivity (square root of the sum of synapses). Nodes are coloured by cell class (SN - orange; IN - magenta; MN - blue; effector - grey). (B) Table of cell type and network statistics. (C) Histogram of edge weights. (D) Size of the largest network after removing edges of increasing weights. (E) Number of sensory neurons with different path distances from effector cells. (F) Number of SN, IN, MN and effector cell types in the cell-type-level connectome shown for all nodes, source nodes, sink nodes and cut nodes. (G) Histogram of weighted degree of nodes, plotted for each cell class. (H) Histogram of pagerank of nodes, plotted for each cell class. (I) Weighted degree of nodes in the cell-type connectome in relation to node pagerank-centrality. <br> Figure 4—source data 1. Source data of the cell type connectivity graph in tbl_graph format saved as an R binary file. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure4.png")
```

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Figure4---figure supplement 5). For a graph-based visualisation of information flow, we arranged the nodes according to their relative ration of incoming and outgoing synapses (Figure 4---figure supplement 6). This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

```{r, fig.cap="**Figure 5. Transmitter phenotypes mapped at single-cell-resolution. ** (A) The cell-type connectome graph with nodes coloured based on neurotransmitter phenotype. (B, C) Morphological rendering of neurons with immunogold labelling for pigment dispersing factor (PDF) neuropeptide, ventral (B) and anterior (C) views. (D) Neurons with immunogold labelling for leucokinin. (E) Neurons with immunogold labelling for allatotropin/orexin. (F) Neurons with immunogold labelling for FVamide. (G) Neurons with immunohistochemically mapped expression of sNPF/RYamide, RGWamide and pedal peptide 2/MLD neuropeptides. (H) Neurons with immunohistochemically mapped expression of achatin and immunogold- or immunohistochemically-mapped (SN-IRP2-FMRF) FMRFamide neuropeptide. (I) Neurons with immunogold (SNMIP-vc) or immunohistochemically mapped expression of myoinhibitory peptide (MIP) expression. (J) Neurons with immunohistochemically mapped expression of serotonin and genetically mapped expression of tryptophan hydroxylase (TrpH), a serotonergic marker. (K) Neurons with genetically mapped expression of vesicular glutamate transporter (VGluT), a glutamatergic marker. (L, M) Neurons with genetically mapped expression of vesicular acetylcholine transporter (VAChT) and choline acetyltransferase (ChAT), cholinergic markers, ventral (L) and anterior (M) views. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure5.png")
```

@@

```{r, fig.cap="**Figure 6. Anatomy of head neuropils and global organisation of brain connectivity** (A) Morphological rendering of all head neurons. (B) Head sensory neurons coloured by head sensory ganglia. Abbreviations: DLSO, dorso-lateral sense organ; DSO, dorsal sense organ; SNMB, mushroom body sensory neuron; INMB, mushroom body interneuron; INMBintr, mushroom-body-intrinsic interneuron; AO SN, apical organ sensory neuron; AO IN, apical organ interneuron; INcentr, central brain interneuron; vMN, ventral motoneuron. (C) Same cells as in (B) rendered without the cell soma to show neuropil organisation. (D) Head interneurons coloured by head ganglia. (E) Same cells as in (D) rendered without the cell soma to show neuropil organisation. (F) Rendering of other sensory cell types that do not form separate neuropils. (G) Rendering of head motoneurons. (H) Sankey circuit diagram showing information flow (from left to right) based on synaptic connectivity between cell categories of head ganglia. Bars represent groups of neurons, grey connecting lines represent synaptic connections (pre-to-post organised left-to-right). Magenta lines represent right-to-left connections. Only connections with >10 synapses are shown. <br> Figure 6—source data 1. Connectivity matrix of the network in panel H. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure6.png")
```

@@

```{r, fig.cap="**Figure 7. Brain circuits and cell types** (A) Schematic diagram of the postural control system of vMN neurons with their direct sensory neuron inputs and motor outputs. (B-C) Morphological rendering of neurons and effector cells of the postural control system, ventral (B) and anterior (C) views. (D) Sankey diagram of synaptic connectivity in the postural control system. (E) Sankey diagram of synaptic connectivity of MNant motoneurons. (F) Morphological rendering of the MNant ciliomotor circuit, anterior view. (G) Morphological rendering of eyespot and visual eye photoreceptors and their direct postsynaptic partners. (H) Sankey diagram of synaptic connectivity of the eyespot and visual eyes. (I) Histogram of the number of cells per head neuronal cell type. (J) Number of presynaptic neuron types for different head cell types (at least 4 synapses). (K)  Number of postsynaptic cell types for different head neuronal cell types (>3 synapses). In B, C, F, G the outline of the yolk is shown for reference. <br> Figure 7—source data 1. Connectivity matrix of the network in panel D. <br>Figure 7—source data 2. Connectivity matrix of the network in panel E. <br> Figure 7—source data 3. Connectivity matrix of the network in panel H. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure7.png")
```

@@

```{r, fig.cap="**Figure 8. Anatomy of the *Platynereis* mushroom bodies** (A) Morphological rendering of the *Platynereis* mushroom bodies in the context of the entire nervous system (neurites shown in cyan), ventral view. (B-G) Morphological renderings of all mushroom body sensory neurons (B), intrinsic interneurons (C), output neurons (D), and all neurons in three different views (E-G). The visual eyes, yolk outline and the stomodeum are shown in grey for reference. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure8.png")
```

@@

The outputs of the mushroom body circuits also include projection neurons of the central brain, including the INW, INlasso, INproT2 and INdecussHook neurons (Figure 9H-Q).
Among these the INproT2 neurons are premotor neurons synapsing on trunk MNring motoneurons, indicating a shallow sensory-motor organisation of mushroom body outputs.

@@

```{r, fig.cap="**Figure 9. Parallel circuit organisation of the mushroom bodies** (A) Connectivity of the mushroom body by neuron category with morphological rendering for each category shown below. (B) Connectivity of mushroom-body neurons and their outputs. Nodes represent neurons grouped by cell type, arrows show synaptic connectivity. (C-D) Sensory inputs to the mushroom bodies (other than SNMB), anterior (C) and ventral (D) views. (E) Palp sensory neurons and their interneuron targets in the mushroom body. (F) Antennal and SNbronto sensory neurons and their interneuron targets in the mushroom body. (G) SNhorn and SNstiff sensory neurons and their interneuron targets in the mushroom body. (H) Interneuron inputs to the mushroom body. (I-L) Inputs to INMBtype2 (I), INMBtype5 (J), INMBtype6 (K) and INMBtype7 (L) mushroom body interneurons. (M) INMBtype9 and its postsynaptic target, the INproT2 premotor projection neuron. (N-O) Pre- and postsynaptic partners of INproT2, anterior (N) and ventral (O) views. (P) Partner of the INW projection neurons. (Q) Partners of the INdecusshook projection neurons. Yolk outline and all mushroom body neurons are shown in grey for reference. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron; MBSN, mushroom body sensory neuron; MBintrIN, mushroom body intrinsic interneuron; MBON, mushroom body output neuron. <br> Figure 9—source data 1. Connectivity matrix of the network in panel A. <br> Figure 9—source data 2. Connectivity matrix of the network in panel B. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure9.png")
```

@@

The trunk-projecting brain neurons have either a decussating (crossing the midline in the head and then descending) or ipsilaterally descending morphology.
The two types project to different domains of the VNC and have different target neurons (Figure 10---figure supplement 1E, F).
Decussating and descending neurons receive most of their incoming synapses in the central head neuropil and form most of their outgoing synapses in the VNC (Figure 10---figure supplement 1C, D, H-M).

@@

```{r, fig.cap="**Figure 10. Connections between the head and the trunk. ** (A) Morphological rendering of head (cyan) and trunk (red) cells, which are part of the connectome. (B) Connectome graph with head (cyan) and trunk (red) cells coloured separately. (C) Morphological rendering of all head neurons with descending projections into the ventral nerve cord. (D) Morphological rendering of all trunk neurons with ascending projections into the head. (E) Morphological rendering of all synaptic targets of head neurons in the trunk coloured by cell class. (F) Distribution across classes of trunk targets of head neurons ordered by the number of head to trunk synapses (top 50 neurons shown). (G) Morphological rendering of all synaptic targets of trunk neurons in the head coloured by cell class. (H) Distribution across classes of head targets of trunk neurons ordered by the number of trunk to head synapses (top 50 neurons shown). <br> Figure 10—source data 1. Source data for panel F. <br> Figure 10—source data 2. Source data for panel H. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure10.png")
```

@@

```{r, fig.cap="**Figure 11. Intersegmental connectivity in the larva ** (A) All cells in the body, coloured by body region. (B) Connectome graph with nodes coloured by segment. (C) Sankey connectivity diagram of sensory (orange), interneurons (cyan), motoneurons (blue) and effectors (purple) in the different body regions. Edge thickness is proportional to the number of synaptic connections. (D) Number of synaptic connections linking the six body regions. (E) (F) Distribution of synapses across different target cell classes for every source cell class (SN, IN, MN, grouped by body region). (G) Distribution of synapses across target body regions for every source cell class (SN, IN, MN, grouped by body region). (H-K) Morphological rendering of all neuronal cell types in segments 1-3 and the pygidium showing cross-segmental neurite projections. (L-N) Neurons with global (whole-body) projections in the head (L), first segment (M) and the pygidium (N). In (H-N) the yolk outline is shown in grey for reference. <br> Figure 11—source data 1. Source data for panel B and C.<br>Figure 11—source data 2. Source data for panel D.<br>Figure 11—source data 3. Source data for panel E.<br>Figure 11—source data 4. Source data for panel F and G. \n\n"}
#| echo: false
knitr::include_graphics("Figures/Figure11.png")
```

@@

Ranking trunk and pygidialmaps cell types by various network centrality measures often identified segment-specific cells as strongly connected (e.g. pygPBunp, Ser-tr1, Loop)(Figure 12---figure supplement 2).

>>>>

---
title: "Whole-body connectome of a segmented annelid larva"
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---
\

Gáspár Jékely^1,2,\*^ , Sanja Jasek^1,2^, Martin Gühmann^3^, Luis Alberto Bezares Calderón^2^, Elizabeth A. Williams^4^, Réza Shahidi^5^\

^1^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
^2^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
^3^School of Biological Sciences, University of Bristol, Bristol, United Kingdom\
^4^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\
^5^Electron Microscopy Core Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany\
^\*^Correspondence: gaspar.jekely\@cos.uni-heidelberg.de

@@

We previously reported synaptic connectomes for several whole-body circuits from a transmission EM  volume of a three-day-old larva [@Randel2015].
These include the visual, startle, ciliomotor, nuchal organ, and neurosecretory systems [@bezares-calderon2018; @Randel2015; @shahidi2015serial; @veraszto2017; @veraszto2018; @williams2017].
Here, we report the complete synaptic connectome and the cell-type complement from the same volume of a three-day-old *Platynereis* larva.
Recently we also reported the desmosomal connectome and motoneuron innervation of the somatic musculature [@jasek2022] and a preprint of a preliminary analysis of the synaptic connectome [@veraszto2020] that this work supersedes.
The full connectome reconstruction has now allowed us to uncover several new circuits and to consider all circuits in a whole-body context.
We also found several neurons that span the entire length of the larva, highlighting the strength of a whole-body dataset.
These cells and their circuits allow us to generate hypotheses on how whole-body coordination may be achieved by the larval nervous system.
The connectome also allows us to address long-standing hypotheses about the origin of the segmented annelid body-plan and explore patterns of circuit evolution.

@@

```{r, out.width = "100%", fig.cap="**Figure 1. Overview of nervous-system anatomy in the three-day-old *Platynereis* larva.** (A) Stylised scanning electron microscopy image of a three-day-old segmented *Platynereis* larva. (B) Morphological rendering of all cells in the EM volume. Spheres represent the position of cell somas. (C) Neurite processes of all neurons in the larva coloured by neuron type (ventral view). Soma are not shown. (D) All effector cells in the larva, including ciliated cells (yellow), glands (grey) and muscles (red). Spheres indicate cell soma (not shown for muscle cells) (E) Summary of cell numbers of different categories in the larva.  (F-H) All presynaptic sites coloured by neuron type shown in (F) ventral, lateral (G), and frontal (H) views. (I) Morphological rendering of all sensory and interneurons on the left side of the larva.  (J) Morphological rendering of all sensory and motor neurons on the left side of the larva. Horizontal lines indicate the position of the cross-sections in (K). (K) Cross-section view of sensory, inter- and motor neuron projections in the ventral nerve cord neuropil. Cross-sections at three antero-posterior positions are shown. Numbers indicate the position of the cross-sections as marked in (J). (L) Position of presynaptic and postsynaptic sites on all sensory neurons of the connectome. (M) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in sensory neurons. (N) Position of presynaptic and postsynaptic sites on all interneurons of the connectome. (O) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in interneurons. (P) Position of presynaptic and postsynaptic sites on all motor neurons of the connectome. (Q) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in motor neurons. In all anatomical renderings, the outline of the yolk is shown in grey. In (G), the body outline is also shown. Aciculae and chaetae are also shown in grey as segmental markers. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron, sg0-3, segments 0-3. <br> Figure 1—source data 1. Source data for panels M, O, Q. " }
#| echo: false

knitr::include_graphics("Figures/Figure1.png")
```

@@

To identify more strongly connected subgraphs within the connectome, we used community detection with the Leiden algorithm [@traag2019louvain].
This analysis combined with force-field-based clustering highlighted several modules (Figures 2).
Nodes in a module are more strongly interconnected than to nodes in other modules (Figure 2---figure supplement 3).
We named the modules based on their primary effector organs or other dominant anatomical characters. We would like to note that this analysis can recover varying numbers of modules depending on the resolution parameter (Figure 2—figure supplement 2C). The selected resolution parameter subdivides the network into communities that we think best represent functional units.

@@

The modules are also interconnected among themselves suggesting crosstalk. Most sensory neurons occur in the visual, the two mechanosensory, the mushroom body and the anterior neurosecretory modules (Figure 2---figure supplement 1J). In a Sankey network diagram representing information flow, these modules are upstream of the motor-dominated pigmentmotor, ciliomotor, and musclemotor modules (Figure 2---figure supplement 2B).

```{r, out.width = "100%", fig.cap="**Figure 2. Modularity of the *Platynereis* larval connectome.** Full connectome graph of the *P. dumerilii* three-day-old larva chemical synapse connectome. Nodes represent individual cells, edges represent synaptic connectivity. Nodes are coloured by modules. Node sizes are proportional to weighted degree. The individual panels show the morphology of all cells in each module with the outline of the yolk and aciculae shown in grey. <br> Figure 2—source data 1. Source data of the connectome graph in tibble graph (tbl_graph) format saved as an R binary object. "}
#| echo: false 
knitr::include_graphics("Figures/Figure2.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 3. Cell-type complement of the three-day-old *Platynereis* larva.** Segmental distribution and number of sensory neuron types (A), interneuron types (B) and motoneuron types (C). Histogram of the number of cells per neuronal (D) and non-neuronal (E) cell types. (F) Segmental distribution and number of non-neuronal cell types. The number in each box refers to the number of cells of the indicated cell type in the indicated segment. <br> Figure 3—source data 1. Source data for panels A-C. <br> Figure 3—source data 2. Source data for panel F. "}
#| echo: false
knitr::include_graphics("Figures/Figure3.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 4. Cell-type-level connectivity of the *Platynereis* larval connectome.** (A) The cell-type connectome of the three-day-old larva. Nodes represent grouped cells of the same type, edges represent synaptic connectivity (square root of the sum of synapses). Nodes are coloured by cell class (SN - orange; IN - magenta; MN - blue; effector - grey). (B) Table of cell type and network statistics. (C) Histogram of edge weights. (D) Size of the largest network after removing edges of increasing weights. (E) Number of sensory neurons with different path distances from effector cells. (F) Number of SN, IN, MN and effector cell types in the cell-type-level connectome shown for all nodes, source nodes, sink nodes and cut nodes. (G) Histogram of weighted degree of nodes, plotted for each cell class. (H) Histogram of pagerank of nodes, plotted for each cell class. (I) Weighted degree of nodes in the cell-type connectome in relation to node pagerank-centrality. <br> Figure 4—source data 1. Source data of the cell type connectivity graph in tbl_graph format saved as an R binary file. "}
#| echo: false
knitr::include_graphics("Figures/Figure4.png")
```

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Figure 4---figure supplement 5). For a graph-based visualisation of information flow, we arranged the nodes according to their relative ratio of incoming and outgoing synapses (Figure 4---figure supplement 6). This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

```{r, out.width = "70%", fig.cap="**Figure 5. Transmitter phenotypes mapped at single-cell-resolution. ** (A) The cell-type connectome graph with nodes coloured based on neurotransmitter phenotype. (B, C) Morphological rendering of neurons with immunogold labelling for pigment dispersing factor (PDF) neuropeptide, ventral (B) and anterior (C) views. (D) Neurons with immunogold labelling for leucokinin. (E) Neurons with immunogold labelling for allatotropin/orexin. (F) Neurons with immunogold labelling for FVamide. (G) Neurons with immunohistochemically mapped expression of sNPF/RYamide, RGWamide and pedal peptide 2/MLD neuropeptides. (H) Neurons with immunohistochemically mapped expression of achatin and immunogold- or immunohistochemically-mapped (SN-IRP2-FMRF) FMRFamide neuropeptide. (I) Neurons with immunogold (SNMIP-vc) or immunohistochemically mapped expression of myoinhibitory peptide (MIP) expression. (J) Neurons with immunohistochemically mapped expression of serotonin and genetically mapped expression of tryptophan hydroxylase (TrpH), a serotonergic marker. (K) Neurons with genetically mapped expression of vesicular glutamate transporter (VGluT), a glutamatergic marker. (L, M) Neurons with genetically mapped expression of vesicular acetylcholine transporter (VAChT) and choline acetyltransferase (ChAT), cholinergic markers, ventral (L) and anterior (M) views. "}
#| echo: false
knitr::include_graphics("Figures/Figure5.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 6. Anatomy of head neuropils and global organisation of brain connectivity** (A) Morphological rendering of all head neurons. (B) Head sensory neurons coloured by head sensory ganglia. Abbreviations: DLSO, dorso-lateral sense organ; DSO, dorsal sense organ; SNMB, mushroom body sensory neuron; INMB, mushroom body interneuron; INMBintr, mushroom-body-intrinsic interneuron; AO SN, apical organ sensory neuron; AO IN, apical organ interneuron; INcentr, central brain interneuron; vMN, ventral motoneuron. (C) Same cells as in (B) rendered without the cell soma to show neuropil organisation. (D) Head interneurons coloured by head ganglia. (E) Same cells as in (D) rendered without the cell soma to show neuropil organisation. (F) Rendering of other sensory cell types that do not form separate neuropils. (G) Rendering of head motoneurons. (H) Sankey circuit diagram showing information flow (from left to right) based on synaptic connectivity between cell categories of head ganglia. Bars represent groups of neurons, grey connecting lines represent synaptic connections (pre-to-post organised left-to-right). Magenta lines represent right-to-left connections. Only connections with >10 synapses are shown. <br> Figure 6—source data 1. Connectivity matrix of the network in panel H. "}
#| echo: false
knitr::include_graphics("Figures/Figure6.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 7. Brain circuits and cell types** (A) Schematic diagram of the postural control system of vMN neurons with their direct sensory neuron inputs and motor outputs. (B-C) Morphological rendering of neurons and effector cells of the postural control system, ventral (B) and anterior (C) views. (D) Sankey diagram of synaptic connectivity in the postural control system. (E) Sankey diagram of synaptic connectivity of MNant motoneurons. (F) Morphological rendering of the MNant ciliomotor circuit, anterior view. (G) Morphological rendering of eyespot and visual eye photoreceptors and their direct postsynaptic partners. (H) Sankey diagram of synaptic connectivity of the eyespot and visual eyes. (I) Histogram of the number of cells per head neuronal cell type. (J) Number of presynaptic neuron types for different head cell types (at least 4 synapses). (K)  Number of postsynaptic cell types for different head neuronal cell types (>3 synapses). In B, C, F, G the outline of the yolk is shown for reference. <br> Figure 7—source data 1. Connectivity matrix of the network in panel D. <br>Figure 7—source data 2. Connectivity matrix of the network in panel E. <br> Figure 7—source data 3. Connectivity matrix of the network in panel H. "}
#| echo: false
knitr::include_graphics("Figures/Figure7.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 8. Anatomy of the *Platynereis* mushroom bodies** (A) Morphological rendering of the *Platynereis* mushroom bodies in the context of the entire nervous system (neurites shown in cyan), ventral view. (B-G) Morphological renderings of all mushroom body sensory neurons (B), intrinsic interneurons (C), output neurons (D), and all neurons in three different views (E-G). The visual eyes, yolk outline and the stomodeum are shown in grey for reference. "}
#| echo: false
knitr::include_graphics("Figures/Figure8.png")
```

@@

The outputs of the mushroom body circuits also include projection neurons of the central brain, including the INW, INlasso, INproT2 and INdecussHook neurons (Figure 9H-Q).
Among these the INproT2 neurons are premotor neurons synapsing on trunk MNring motoneurons, indicating a shallow sensory-motor organisation of mushroom-body outputs.

@@

```{r, out.width = "100%", fig.cap="**Figure 9. Parallel circuit organisation of the mushroom bodies** (A) Connectivity of the mushroom body by neuron category with morphological rendering for each category shown below. (B) Connectivity of mushroom-body neurons and their outputs. Nodes represent neurons grouped by cell type, arrows show synaptic connectivity. (C-D) Sensory inputs to the mushroom bodies (other than SNMB), anterior (C) and ventral (D) views. (E) Palp sensory neurons and their interneuron targets in the mushroom body. (F) Antennal and SNbronto sensory neurons and their interneuron targets in the mushroom body. (G) SNhorn and SNstiff sensory neurons and their interneuron targets in the mushroom body. (H) Interneuron inputs to the mushroom body. (I-L) Inputs to INMBtype2 (I), INMBtype5 (J), INMBtype6 (K) and INMBtype7 (L) mushroom body interneurons. (M) INMBtype9 and its postsynaptic target, the INproT2 premotor projection neuron. (N-O) Pre- and postsynaptic partners of INproT2, anterior (N) and ventral (O) views. (P) Partner of the INW projection neurons. (Q) Partners of the INdecusshook projection neurons. Yolk outline and all mushroom body neurons are shown in grey for reference. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron; MBSN, mushroom body sensory neuron; MBintrIN, mushroom body intrinsic interneuron; MBON, mushroom body output neuron. <br> Figure 9—source data 1. Connectivity matrix of the network in panel A. <br> Figure 9—source data 2. Connectivity matrix of the network in panel B. "}
#| echo: false
knitr::include_graphics("Figures/Figure9.png")
```

@@

The trunk-projecting brain neurons have either a decussating or ipsilaterally descending morphology.
The two types project to different domains of the VNC and have different target neurons (Figure 10---figure supplement 1E, F).
Decussating and descending neurons receive most of their incoming synapses in the central head neuropil and form most of their outgoing synapses in the VNC (Figure 10---figure supplement 1C, D, H-M).

@@

```{r, out.width = "100%", fig.cap="**Figure 10. Connections between the head and the trunk. ** (A) Morphological rendering of head (cyan) and trunk (red) cells, which are part of the connectome. (B) Connectome graph with head (cyan) and trunk (red) cells coloured separately. (C) Morphological rendering of all head neurons with descending projections into the ventral nerve cord. (D) Morphological rendering of all trunk neurons with ascending projections into the head. (E) Morphological rendering of all synaptic targets of head neurons in the trunk coloured by cell class. (F) Distribution across classes of trunk targets of head neurons ordered by the number of head to trunk synapses (top 50 neurons shown). (G) Morphological rendering of all synaptic targets of trunk neurons in the head coloured by cell class. (H) Distribution across classes of head targets of trunk neurons ordered by the number of trunk to head synapses (top 50 neurons shown). <br> Figure 10—source data 1. Source data for panel F. <br> Figure 10—source data 2. Source data for panel H. "}
#| echo: false
knitr::include_graphics("Figures/Figure10.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 11. Intersegmental connectivity in the larva ** (A) All cells in the body, coloured by body region. (B) Connectome graph with nodes coloured by segment. (C) Sankey connectivity diagram of sensory (orange), interneurons (cyan), motoneurons (blue) and effectors (purple) in the different body regions. Edge thickness is proportional to the number of synaptic connections. (D) Number of synaptic connections linking the six body regions. (E) (F) Distribution of synapses across different target cell classes for every source cell class (SN, IN, MN, grouped by body region). (G) Distribution of synapses across target body regions for every source cell class (SN, IN, MN, grouped by body region). (H-K) Morphological rendering of all neuronal cell types in segments 1-3 and the pygidium showing cross-segmental neurite projections. (L-N) Neurons with global (whole-body) projections in the head (L), first segment (M) and the pygidium (N). In (H-N) the yolk outline is shown in grey for reference. <br> Figure 11—source data 1. Source data for panel B and C.<br>Figure 11—source data 2. Source data for panel D.<br>Figure 11—source data 3. Source data for panel E.<br>Figure 11—source data 4. Source data for panel F and G. "}
#| echo: false
knitr::include_graphics("Figures/Figure11.png")
```

@@

Ranking trunk and pygidial cell types by various network centrality measures often identified segment-specific cells as strongly connected (e.g. pygPBunp, Ser-tr1, Loop)(Figure 12---figure supplement 2).
